---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fastqindexr

<!-- badges: start -->
[![R-CMD-check](https://github.com/brendanf/fastqindexr/actions/workflows/r-cmd-check.yaml/badge.svg)](https://github.com/brendanf/fastqindexr/actions/workflows/r-cmd-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/brendanf/fastqindexr/graph/badge.svg)](https://app.codecov.io/gh/brendanf/fastqindexr)
<!-- badges: end -->


`fastqindexr` builds an in-memory index over one or more FASTA/FASTQ files
(gzip-compressed or plain), then extracts records by numerical index without
scanning from the beginning each time.
The package wraps an Rcpp bridge around adapted open-source
[FastqIndEx](https://github.com/DKFZ-ODCF/FastqIndEx) C++ core components
in order to seek to specific records in a gzipped file without decompressing the
entire file.
It can also read binary `.fqi` indexes produced by the FastqIndEx CLI.
It is not written or maintained by the authors of FastqIndEx.

Additional features implemented beyond the capabilities of FastqIndEx:

  - Fast extraction of arbitrary subsets of records, rather than only
    contiguous ranges.
  - Extraction of records in any order, including duplicates.
  - Multiple file support, where the files are treated as one logical
    concatenated record stream.
  - Support for FASTA files with multiple sequence lines per record.
  - Support for extracting different subsets of records to multiple outputs
    (files or R objects) in a single query and a single read through the source
    file(s).
  - Support for indexing and extracting records from non-gzipped files using
    the same API.
  - Support for extracting records from source file(s) without building an
    index first, with only a single scan through the source file(s).
    This option is useful for one-off extraction where the index build cost is
    not justified.

Many of these features are also implemented using the
[`Biostrings`](https://bioconductor.org/packages/Biostrings/) package's
`fasta.index()` function.
Additional features implemented beyond the capabilities of `Biostrings` are:

  - Support for indexing and extraction from FASTQ files.
  - Support for extracting records to multiple outputs in a single query.
  - Support for extracting non-contiguous records from a non-indexed source file
    without reading the entire file in memory.
  - Greatly improved performance for non-contiguous extraction from gzipped
    files.

Additional features which are present in `Biostrings` but not implemented in
`fastqindexr` are:

  - Extraction of records by text sequence ID or length (or both) rather than
    numerical index.
  - Extracting subsequences rather than entire records.

## Installation

You can install the development version of fastqindexr from GitHub with:

``` r
# install.packages("pak")
pak::pak("brendanf/fastqindexr")
```

## Example

Build an index once, then request records in any order (including duplicates):

```{r example}
library(fastqindexr)

# Make a tiny gzipped FASTQ for demonstration.
path <- tempfile(fileext = ".fastq.gz")
con <- gzfile(path, "wt")
writeLines(
  c(
    "@read1", "ACGT", "+", "!!!!",
    "@read2", "TTAA", "+", "####",
    "@read3", "GGCC", "+", "$$$$"
  ),
  con
)
close(con)

# Index the file (type can also be "auto" or "fasta").
idx <- create_index(path, type = "fastq")
idx

# Extract by 1-based record ID, preserving request order.
extract_sequences(idx, seq_idx = c(3, 1, 1))

# Or write the same selection to a FASTA/FASTQ file (plain or .gz).
out <- tempfile(fileext = ".fastq")
extract_sequences_to_file(idx, seq_idx = c(3, 1, 1), outfile = out)
unlink(c(path, out))
```

## Benchmarking

Here is a benchmark comparing extraction of sequences from a file using
`Biostrings` vs `fastqindexr`, for different access
patterns and compression modes.

```{r benchmark}
# Install Biostrings if not already installed.
# install.packages("BiocManager")
# BiocManager::install("Biostrings")
library(Biostrings)
library(fastqindexr)

set.seed(1)
n_records <- 10000L
n_extract <- 2000L

# Three different request patterns in increasing order of difficulty:
# 1) A contiguous range of 2000 records, but not at the beginning;
# 2) 2000 unique records, randomly scattered through the file but in increasing
#    order;
# 3) 2000 randomly selected records, including potential duplicates, in
#    arbitrary order.

requests <- list(
  contiguous = (n_records %/% 2) + seq_len(n_extract),
  sorted_unique = sort(sample.int(n_records, n_extract, replace = FALSE)),
  unsorted_dup = sample.int(n_records, n_extract, replace = TRUE)
)

benchmarks <- data.frame(
  compression = character(0),
  implementation = character(0),
  operation = character(0),
  elapsed = numeric(0)
)

# loop over compression modes (gzip / plain)
for (comp in c("gzip", "plain")) {
  if (comp == "gzip") {
    fileext <- ".fa.gz"
  } else {
    fileext <- ".fa"
  }
  tmp <- tempfile(fileext = fileext)
  make_benchmark_fasta(tmp, n = n_records, width = 120)

  # loop over implementations (fastqindexr / Biostrings)
  for (impl in c("fastqindexr", "Biostrings")) {
    index_time <- system.time(
      if (impl == "fastqindexr") {
        idx <- create_index(tmp, type = "fasta")
      } else {
        idx <- Biostrings::fasta.index(tmp, seqtype = "DNA")
      }
    )

    benchmarks <- rbind(
      benchmarks,
      data.frame(
        compression = comp,
        implementation = impl,
        operation = "index",
        elapsed = index_time[["elapsed"]]
      )
    )

    # loop over request patterns
    for (r in names(requests)) {
      ids <- requests[[r]]
      extract_time <- system.time(
        if (impl == "fastqindexr") {
          res <- extract_sequences(idx, ids)
        } else {
          res <- Biostrings::readDNAStringSet(idx[ids, ])
        }
      )
      benchmarks <- rbind(
        benchmarks,
        data.frame(
          compression = comp,
          implementation = impl,
          operation = r,
          elapsed = extract_time[["elapsed"]]
        )
      )
    }
  }
  unlink(tmp)
}

benchmarks
```

Although `fastqindexr` supports extracting from both gzipped and plain
FASTA/FASTQ files, it is only substantially faster than `Biostrings`
in the cases where `Biostrings` struggles the most: when extracting
*non-contiguous* sets from *gzipped* files.  Note the logarithmic scale!

```{r benchmark-plot, dev="png", fig.width = 8, fig.height = 6}
library(ggplot2)
benchmarks$operation <- factor(
  benchmarks$operation,
  levels = c("unsorted_dup", "sorted_unique", "contiguous", "index")
)
ggplot(
  benchmarks,
  aes(y = operation, x = elapsed * 1000, fill = implementation)
) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ compression, ncol = 1) +
  scale_x_log10(name = "Elapsed (ms)") +
  scale_y_discrete(name = "Operation") +
  theme_minimal() +
  theme(legend.position = "bottom")
```


## API at a glance

- `create_index(files, type = c("auto", "fasta", "fastq"))`
  - accepts one or many existing FASTA/FASTQ files (all gzip or all plain)
  - returns a `fastqindexr_index` object, with subclass
    `fastqindexr_gzip_index` or `fastqindexr_plain_index`
  - this object can be saved to disk with `saveRDS()` and loaded in a different
    session with `readRDS()` (or serialized/deserialized in other ways, such as
    with the [`qs2`](https://cran.r-project.org/package=qs2) package)
- `read_fqi_index(fqi_path, files = NULL, type = c("auto", "fasta", "fastq"))`
  - turns one or more FastqIndEx `.fqi` files into a `fastqindexr_index` for use
    with the extract functions
- `extract_sequences(index, seq_idx, file = NULL, return = c("data.frame", "list", "seq"))`
  - `seq_idx` are 1-based positive integer record IDs
  - returns rows in the same order as `seq_idx`
  - duplicate IDs are allowed and duplicated in output
  - for FASTQ, output includes `seq_id`, `seq`, and `qual` when `return` is
    `"data.frame"` or `"list"`; for FASTA, it includes `seq_id` and `seq`
  - with `return = "list"`, a list with the same fields; with `return = "seq"`,
    sequences only, named by `seq_id` (duplicates in names are allowed; quality
    lines are not read from the source in that mode for FASTQ)
- `extract_sequences_to_file(index, seq_idx, file = NULL, outfile, ...)`
  - writes records in order to a plain or gzip output file (default compression
    follows the `outfile` name)
  - preserves request order and duplicate IDs exactly; supports
    `type = "auto"`, `"fasta"`, or `"fastq"` output
  - when the request is strictly increasing and `append` is `FALSE`, extraction
    does not need to build a full in-memory map of every unique ID; with
    `append` or a non-sorted request, the implementation may buffer all unique
    records like `extract_sequences()`
- `make_benchmark_fasta(path, n = 5000, width = 80)`
  - writes synthetic gzipped FASTA for repeatable benchmark setup
- `make_benchmark_fastq(path, n = 5000, width = 80)`
  - writes synthetic gzipped FASTQ for repeatable benchmark setup

## Important format assumptions

- FASTA records are header-driven (`>` starts a record); sequence may span
  multiple lines
- files are treated as a logical concatenated stream in the order supplied to
  `create_index()`
- if files move after indexing, you can provide replacement paths via the
  `file` argument in `extract_sequences()` and `extract_sequences_to_file()`
